Ownership types in theory and practice (in Rust)

### 2023-01-18 paper/paper.tex
Fleshed out owners-as-dominators section

The core concept of ownership types is that objects can be \emph{owners} of other objects.
Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{outside} \emph{b} if \emph{b} is the same object as \emph{a} or \emph{b} is \emph{owned} by \emph{a}, transitively.
The \emph{outside} relation is the converse of the \emph{inside} relation.

@@

Yang Zhao and John Boyland \cite{permission-ownership-types} specify that, with owners-as-dominators, each regular object is owned by exactly one object and can be the owner of zero or more objects.
Objects can also be owned by a special global object called \emph{world}, which cannot be owned anymore.
This \emph{world} object is the root of the acyclic ownership hierarchy tree that represents the ownership relation.
In this tree structure, each object is inside its owner.
This model is named \emph{owners-as-dominators} because any object acting as an owner is considered to be a \emph{dominator} for all of its owned objects.

@@

    \draw[->,thick,dashed,green] (Car) edge[bend left] node[right] {owner} (Engine);
    % \draw[->,thick,dashed,green] (Engine) edge[bend left] node[left] {inside} (Car);
    \draw[->,thick,dashed,green] (Engine) edge[bend left=25] node[left] {inside} (World);
    \draw[->,thick,dashed,green] (Car) edge node[above] {sibling} (Driver);
    \draw[->,thick,dashed,red] (World) edge[bend right=90] node[left] {invalid} (Engine);
    \draw[->,thick,dashed,red] (Driver) edge[bend left] node[right] {invalid} (Engine);
  \end{tikzpicture}

@@

Figure \ref{fig:owners-as-dominators} was adapted from an example by Clarke et al. \cite{flexible-alias-protection} and demonstrates valid (green) and an invalid (red) references in an ownership hierarchy.
% TODO: make more precise
The validity of each valid reference is proven by each of the rules defined above.

\begin{enumerate}
  \item The object \emph{car} owns the object \emph{engine}, meaning that the former can validly reference the latter according to the first rule.
  \item The objects \emph{car} and \emph{driver} are siblings, meaning that they can validly reference each other according to the second rule.
  \item The object \emph{world} is outside the object \emph{engine} because \emph{engine} is transitively owned by \emph{world} via \emph{car}.
  Therefore, according to the third rule, \emph{engine} can validly reference \emph{world}.
\end{enumerate}

\paragraph{Implications}

An implication of this model is that any external reference to an object is only possible through its owner.
This is also examplified in Figure \ref{fig:owners-as-dominators} through the object \emph{world} trying to access \emph{engine}.
This is not allowed in the owners-as-dominators model.
Instead, \emph{world} can only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.

Its strictness can be seen as both an advantage and disadvantage of the owners-as-dominators model.
% TODO: reference ownership survey better, as this is nearly copied
The advantage is that model provides a simple, clear and strong guarantee that allows for reasoning about various properties of the code.
% TODO: provide example of advantages
However, it also makes programming more difficult by not allowing common idioms that involve aliasing.
% TODO: example of such idiom that isn't representable

@@

% IMPORTANT TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?

@@

\subsection{Owners-as-ombudsmen}
\label{sec:owners-as-ombudsmen}

% TODO: is it obvious that this contains exact sentences from citation?
The paper that first introduced the idea of \emph{owners-as-ombudsmen} \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
% TODO: this is the same as dominators
With owners-as-dominators, every path from the root to a given object must pass through that object's owner.
Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.
The owners-as-ombudsmen approach, on the other hand, relaxes the single bridge object constraint and allows several bridge objects to collectively define an aggregate with a shared representation.
These bridge objects are called \emph{ombudsmen} to emphasize their benevolent nature.

% The last flavor of ownership types is the owners-as-ombudsmen approach \cite{ownership-types-survey}, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
% This is achieved by allowing multiple objects to define a shared aggregate owner.

% Extending the rules stated for owners-as-dominators, \emph{a} can also validly reference \emph{b} if \emph{a} is owned by aggregate owner \emph{b}.
For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-ombudsmen must adhere to at least one of the following rules \cite{ownership-types-survey}.
\begin{enumerate}
  \item \emph{a} is the owner of \emph{b},
  \item \emph{a} and \emph{b} are siblings, or
  \item \emph{b} is outside of \emph{a}, or
  \item \emph{a} is owned by aggregate owner \emph{b}
\end{enumerate}

\subsection{Advantages and disadvantages}

@@

% TODO: possibly to do

@@

>>>>

The core concept of ownership types is that objects can be \emph{owners} of other objects.
Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
% Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{outside} \emph{b} if \emph{b} is the same object as \emph{a} or \emph{b} is \emph{owned} by \emph{a}, transitively.
The \emph{outside} relation is the converse of the \emph{inside} relation.

@@

Yang Zhao and John Boyland \cite{permission-ownership-types} specify that, with owners-as-dominators, each regular object is owned by exactly one object and can be the owner of zero or more objects.
Objects can also be owned by a special global object called \emph{world}, which cannot be owned anymore.
This \emph{world} object is the root of the acyclic ownership hierarchy tree that represents the ownership relation.
In this tree structure, each object is \emph{inside} its owner.
This model is named \emph{owners-as-dominators} because any object acting as an owner is considered to be a \emph{dominator} for all of its owned objects.

@@

    \draw[->,thick,dashed,green] (Car) edge[bend left] node[right] {owner} (Engine);
    \draw[->,thick,dashed,green] (Engine) edge[bend left] node[left] {inside} (Car);
    \draw[->,thick,dashed,green] (Car) edge node[above] {sibling} (Driver);
    \draw[->,thick,dashed,red] (World) edge[bend right=90] node[left] {invalid} (Engine);
    \draw[->,thick,dashed,red] (Driver) edge[bend left] node[right] {invalid} (Engine);
  \end{tikzpicture}

@@

Figure \ref{fig:owners-as-dominators} was adapted from an example by Clarke et al. \cite{flexible-alias-protection} and demonstrates valid (green) and an invalid (red) references in an ownership hierarchy.
Each of the three reference validity rules that were defined previously will be illustrated using this example figure.

\paragraph{Owner}

The object \emph{car} owns the object \emph{engine}, meaning that \emph{car} can validly reference \emph{engine} according to the first rule.

An owner being able to access its owned objects is the most trivial and intuitive rule.

\paragraph{Outside/inside}

The object \emph{engine} is \emph{inside} the object \emph{car} because \emph{engine} is owned by \emph{car}.
Conversely, \emph{car} is outside \emph{engine}.
Therefore, \emph{engine} can validly reference \emph{car} according to the third rule.

Even the original Flexible Alias Protection model, which these ownership types are based on, imposed the restriction "that the only objects that can access [an object] are the object that owns it and other objects inside that [object]" \cite{flexible-alias-protection}.
The implication is that this model only protects objects from external access.
Internally, an object can still access its owner (and its owner's owner etc.).

\paragraph{Sibling}

The objects \emph{car} and \emph{driver} are siblings, meaning that they can validly reference each other according to the second rule.

The lack of further rules implies that siblings are the only objects that an object can validly reference besides the objects it owns and the objects it is inside of.
The sibling relationship would, therefore, be used for objects that exist independently but can still interact with each other.

\paragraph{Further implications}

Most importantly, these three rules imply that any external reference to an object is only possible through its owner.
This is also examplified in Figure \ref{fig:owners-as-dominators} where the object \emph{world} is trying to access the object \emph{engine}.
This is not allowed in the owners-as-dominators model, as \emph{world} is neither a direct owner, a sibling or inside of \emph{engine}.
Instead, \emph{world} could only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.

Its strictness can be seen as both an advantage and disadvantage of the owners-as-dominators model.
% TODO: reference ownership survey better, as this is nearly copied
The advantage is that the model provides a simple, clear and strong guarantee that allows for reasoning about various properties of the code.
% TODO: provide example of advantages
However, it also makes programming more difficult by not allowing common idioms that involve aliasing.
% TODO: example of such idiom that isn't representable

@@

@@

% \subsection{Owners-as-ombudsmen}
% \label{sec:owners-as-ombudsmen}

% % TODO: is it obvious that this contains exact sentences from citation?
% The paper that first introduced the idea of \emph{owners-as-ombudsmen} \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
% % TODO: this is the same as dominators
% With owners-as-dominators, every path from the root to a given object must pass through that object's owner.
% Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.
% The owners-as-ombudsmen approach, on the other hand, relaxes the single bridge object constraint and allows several bridge objects to collectively define an aggregate with a shared representation.
% These bridge objects are called \emph{ombudsmen} to emphasize their benevolent nature.

% % The last flavor of ownership types is the owners-as-ombudsmen approach \cite{ownership-types-survey}, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
% % This is achieved by allowing multiple objects to define a shared aggregate owner.

% % Extending the rules stated for owners-as-dominators, \emph{a} can also validly reference \emph{b} if \emph{a} is owned by aggregate owner \emph{b}.
% For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-ombudsmen must adhere to at least one of the following rules \cite{ownership-types-survey}.
% \begin{enumerate}
%   \item \emph{a} is the owner of \emph{b},
%   \item \emph{a} and \emph{b} are siblings, or
%   \item \emph{b} is outside of \emph{a}, or
%   \item \emph{a} is owned by aggregate owner \emph{b}
% \end{enumerate}

@@

% TODO: I could easily link this back to rust

@@

% TODO: check if rust implements all rules from dominators or modifiers
% TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?
